Over Twenty Years of Virtual Machine Development and Debugging Through Simulation

### 2018-08-14 paper.tex
keep pass

HERE
Recently, to evaluate new garbage collection algorithm against standard algorithm, we implemented a Mark-Sweep in addition to the existing Mark-Compact for old space garbage collection. The whole implementation was done in the simulator, and only when it was working there, we compiled it to C and it was working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertion, ensuring a specific state is as the VM developer would expect it to be, and stopping the program execution if not. We have basically three levels of assertions. Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed, but they can be performed only in the simulator. Assertions written in Slang are also verified in simulation mode, but they are compiled to C. The C compiler, based on a compilation flag, then picks if it wants to compile the VM with or without assertions. The VM with assertions is used for debugging and to recreate snapshots of runtime juste before they crash. The VM without assertions is the production VM.

\paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator. Since the algorithm was partially written, we could just write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in Smalltalk. In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version. This means that if the algorithm was not working, an assertion would fail or an error such as incorrect memory access would happen in a copy of the heap. The simulator can then reproduce the exact same crash as many times as we want duplicating again and again the same original heap. This is very convenient to debug specific GC bugs because often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state.

>>>>

Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was kind of working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions. Each assertion ensures a specific state is as the VM developer would expect it to be and stops or log the incorrect result if not. We have a multi-level assertion system. Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed easily using high level structures (Sets, etc.), but they can be performed only in the simulator. Assertions written in Slang are executed in the simulator but they are also compiled to C. The C compiler, based on a compilation flag, chooses to compile the VM with or without assertions. The VM with assertions is used for debugging and to recreate snapshots just before a VM crash. The VM without assertions is the production VM.

HERE
\paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator. Since the algorithm was partially written, we had to write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in Smalltalk. In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version. This means that if the algorithm was not working, an assertion would fail or an error such as incorrect memory access would happen in a copy of the heap. The simulator can then reproduce the exact same crash as many times as we want duplicating again and again the same original heap. This is very convenient to debug specific GC bugs because often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state.
